---
title: "A Gapminder exploration"
output: pdf_document
---
```{r package_setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

knitr::opts_chunk$set(
  fig.width = 9,   # increase width
  fig.height = 9,
  out.width = "100%"   # VERY IMPORTANT
)

library(tidyr)
library(dplyr)
library(ggplot2)

source(here::here('notebooks', 'utils.R'))
```


# Data loading
```{r}
t.features = read_csv_locale(here::here('data', 'gapminder-misc.csv'))
t.countries = read_csv_locale(here::here('data', 'gapminder-countries.csv'))

t.all = left_join(t.features, t.countries, by = 'country')

col.features = colnames(t.features)[seq(3, ncol(t.features), by = 2)]
```
Gapminder data were loaded from [Gapminder](https://www.gapminder.org/data/), avalaible (observations and forecasts) from `r length(unique(t.all$country))` countries.
Available features are:

```{r}
t.feature_ranges <- {
  t_values <- t.features %>%
    pivot_longer(
      cols = -c(country, time),
      names_to = "feature",
      values_to = "value"
    ) %>%
    filter(!grepl("_is_forecast$", feature))

  t_flags <- t.features %>%
    pivot_longer(
      cols = ends_with("_is_forecast"),
      names_to = "feature",
      values_to = "is_forecast"
    ) %>%
    mutate(
      feature = sub("_is_forecast$", "", feature),
      is_forecast = is_forecast %in% c(TRUE, "True", "TRUE", "true")
    )

  t_values %>%
    left_join(t_flags, by = c("country", "time", "feature")) %>%
    mutate(is_forecast = coalesce(is_forecast, FALSE)) %>%
    group_by(feature) %>%
    summarise(
      start_observation = min(time[!is.na(value)], na.rm = TRUE),
      start_forecast = if (any(is_forecast)) min(time[is_forecast], na.rm = TRUE) else NA_real_,
      end = max(time[!is.na(value)], na.rm = TRUE),
      .groups = "drop"
    )
}

knitr::kable(
  t.feature_ranges,
  col.names = c('feature', 'recorded from', 'forecasted from', 'until')
)
```
# Starter plots

## One line - one country

```{r}

ggplot(
  data = t.all |> filter(name == 'Switzerland'),
  mapping = aes(
    x = time,
    y = lex,
    # lty = lex_is_forecast
  )
) +
  geom_line(lwd = 1.5) +
  # theme_minimal(base_size = 16) +
  # scale_linetype_discrete(labels = c("Observation", "Forecast")) +
  labs(
    title = 'Switzerland life expectancy over time',
    subtitle = 'Gapminder publishe observations and forecasts',
    x = 'Year',
    y = 'Life expectancy (year)',
    # lty = ''
  )

```

## Mutliple countries

Goal:
 - add one color for each country.
 - the legend for colors must say: "Country"
 - write a caption to tell the story.

```{r}

ggplot(
  data = t.all |> filter(name %in% c('Switzerland', 'Ethiopia')),
  mapping = aes(
    x = time,
    y = lex,
    lty = lex_is_forecast,
    colour= name
  )
) +
  geom_line(lwd = 1.5) +
  theme_minimal(base_size = 16) +
  scale_linetype_discrete(labels = c("Observation", "Forecast")) +
  labs(
    title = 'Switzerland and Ethiopia life expectancy over time',
    subtitle = 'Gapminder published observations and forecasts',
    x = 'Year',
    y = 'Life expectancy (year)',
    lty = '',
    colour = "Country",
    caption = "Evolution of life expectancy in Switzerland and Ethiopia. 
               The Switzerland in the 50's had the Lex of todays Lex in Ethiopia.
               The Lex increases faster in Ethiopia narrowing the gap, 
               the forecast suggests that this will continue"
  )

```

## Child death under 5 vs wealth inequality (Gini coefficient)

Goal:
 - snapshot in 2021
 - x/y with the title metrics
 - circle size with the population
 - color with the region (4 levels)

```{r}

ggplot(
  data = t.all |> filter(time == 2021),
  mapping = aes(
    x = gini,
    y = child_mortality_0_5_year_olds_dying_per_1000_born,
    size= pop,
    colour= income_groups
  )
) +
  geom_point(alpha = 0.7) +
  theme_minimal(base_size = 16) +
  scale_colour_manual(
    values = c(
      "low_income" = "#ee0e0aef",
      "lower_middle_income" = "#fa9600",
      "upper_middle_income" = "#3fd619ef",
      "high_income" = "#0a9aeeef"
    ),
    labels = c(
      "low_income" = "Low income",
      "lower_middle_income" = "Lower middle income",
      "upper_middle_income" = "Upper middle income",
      "high_income" = "High income"
    ),
    breaks = c(
      "low_income",
      "lower_middle_income",
      "upper_middle_income",
      "high_income" 
    )
  ) +
  labs(
    title = 'Child death under 5 vs wealth inequality',
    subtitle = 'Snapshot of year 2021 taken from Gapminder published observations',
    x = 'Gini coefficient',
    y = 'Child death under 5',
    size = 'Population',
    colour = 'Income Groups',
    caption = "The death of childs under 5 years old is poorly related to 
    the gini coefficient neither to the population.
    However, there's a clear relation with the income group the country belongs to."
  )

```